---
title: "Project 2: Zurich Politicians Analysis"
author: "Igor"
date: "2026-09-30"
output: html_document
---
Introduction
In this project, I am analyzing the Zurich politicians dataset (zh_politicians.db). The goal is to load the SQLite database, clean up missing values and date formats, combine the main tables, and do basic exploratory analysis to see trends about gender, political parties, and how long mandates last.

Loading the Data
First, I load the necessary R packages. Then I connect to the SQLite database file in my folder and collect the four main tables (PERSONS, MANDATES, ADDRESSES, and AFFILIATIONS).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Loading packages
library(DBI)
library(RSQLite)
library(tidyverse)
library(lubridate)

# Setting path to database file
db_path <- "zh_politicians.db"

# Check if file exists
if (!file.exists(db_path)) {
  stop("Error: zh_politicians.db not found! Please check file location.")
}

# Connect to SQLite
con <- dbConnect(RSQLite::SQLite(), db_path)

# Download tables into R dataframes
persons_raw      <- tbl(con, "PERSONS") %>% collect()
mandates_raw     <- tbl(con, "MANDATES") %>% collect()
addresses_raw    <- tbl(con, "ADDRESSES") %>% collect()
affiliations_raw <- tbl(con, "AFFILIATIONS") %>% collect()

# Close connection
dbDisconnect(con)
# Loading packages
library(DBI)
library(RSQLite)
library(tidyverse)
library(lubridate)

# Setting path to database file
db_path <- "zh_politicians.db"

# Check if file exists
if (!file.exists(db_path)) {
  stop("Error: zh_politicians.db not found! Please check file location.")
}

# Connect to SQLite
con <- dbConnect(RSQLite::SQLite(), db_path)

# Download tables into R dataframes
persons_raw      <- tbl(con, "PERSONS") %>% collect()
mandates_raw     <- tbl(con, "MANDATES") %>% collect()
addresses_raw    <- tbl(con, "ADDRESSES") %>% collect()
affiliations_raw <- tbl(con, "AFFILIATIONS") %>% collect()

# Close connection
dbDisconnect(con)
```

Data Cleaning and Preparation
Looking at the raw data, I noticed two main issues:

Empty text values "" need to be changed to NA so R knows they are missing values.

The dates for mandate starts and ends are split into separate day, month, and year columns. I will use make_date() from lubridate to make proper date variables.

```{r clean-data}
# Cleaning persons table
persons_clean <- persons_raw %>%
  mutate(
    across(where(is.character), ~na_if(., "")),
    YEAR_OF_DEATH = as.numeric(YEAR_OF_DEATH)
  )

# Cleaning mandates table and creating real dates
mandates_clean <- mandates_raw %>%
  mutate(
    across(where(is.character), ~na_if(., "")),
    START_DATE = make_date(MANDATE_START_YEAR, MANDATE_START_MONTH, MANDATE_START_DAY),
    END_DATE   = make_date(MANDATE_END_YEAR, MANDATE_END_MONTH, MANDATE_END_DAY),
    DURATION_DAYS = as.numeric(END_DATE - START_DATE)
  )

# Cleaning affiliations (parties) table
affiliations_clean <- affiliations_raw %>%
  mutate(
    across(where(is.character), ~na_if(., ""))
  )

# Cleaning addresses table
addresses_clean <- addresses_raw %>%
  mutate(
    across(where(is.character), ~na_if(., ""))
  )
  
```