---
title: "Project 1: Rental Market Data Analysis"
author: "Igor Donchev"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)
```

In this project, I am analyzing a rental dataset scraped from a fake rental agency website. I will scrape the page directly using rvest, clean up column values, convert strings to numbers, create scatterplots and bar charts using ggplot2, run two t-tests, and render an interactive map using leaflet.

---

# Loading Packages

```{r load-packages}
library(tidyverse)
library(rvest)
library(janitor)
library(tidygeocoder)
library(leaflet)
library(knitr)
library(broom)
```

---

# Data Import and Cleaning

## Part 1a: Scraping landing page

Passing a clean URL string directly to `read_html()` to extract all listing cards based on `.col-sm` containers.

```{r scrape-rental-site}
url <- "https://epfl-exts.github.io/rental-scrape/"

page <- read_html(url)
cards <- html_nodes(page, ".col-sm")

rental_raw <- data.frame(
  object_type  = html_text(html_node(cards, ".object-type"), trim = TRUE),
  rooms        = html_text(html_node(cards, ".rooms"), trim = TRUE),
  price        = html_text(html_node(cards, ".price"), trim = TRUE),
  address      = html_text(html_node(cards, ".address"), trim = TRUE),
  living_space = html_text(html_node(cards, ".living-space"), trim = TRUE),
  floor        = html_text(html_node(cards, ".floor"), trim = TRUE),
  availability = html_text(html_node(cards, ".availability"), trim = TRUE),
  stringsAsFactors = FALSE
)
```

## Checking dataset dimensions

```{r check-dim}
dim(rental_raw)
```

---

## Part 1b: Cleaning column names and converting types

Cleaning column names with `clean_names()` and parsing numeric metrics.

```{r clean-rental-data}
rental_clean <- clean_names(rental_raw)

# Extracting numeric values into new columns
rental_clean$price_chf <- parse_number(rental_clean$price)
rental_clean$living_space_m2 <- parse_number(rental_clean$living_space)
rental_clean$floor_num <- parse_number(rental_clean$floor)
rental_clean$rooms_num <- parse_number(rental_clean$rooms)
rental_clean$postcode_cat <- as.character(parse_number(rental_clean$address))

# Identifying address status
address_lower <- str_to_lower(rental_clean$address)
rental_clean$address_status <- ifelse(
  str_detect(address_lower, "on request") | str_detect(address_lower, "sur demande"),
  "On Request",
  "Available"
)

# Calculating price per square meter
rental_clean$price_per_m2 <- rental_clean$price_chf / rental_clean$living_space_m2
```

## Inspecting clean dataset structure

```{r inspect-clean}
glimpse(rental_clean)
```

## Checking numerical ranges

```{r check-ranges}
summary(rental_clean$price_chf)
summary(rental_clean$living_space_m2)
```

---

# Exploratory Data Analysis

## Part 2: Scatterplot of price vs living space

```{r plot-price-space}
ggplot(rental_clean, aes(x = living_space_m2, y = price_chf)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Price vs Living Space",
    subtitle = "Relationship between living space and total rent",
    x = "Living Space (m2)",
    y = "Price (CHF)"
  ) +
  theme_minimal()
```

---

## Part 3: Bar plot of properties by postcode

```{r plot-postcodes}
ggplot(rental_clean, aes(x = postcode_cat, fill = postcode_cat)) +
  geom_bar() +
  labs(
    title = "Number of Properties by Postcode",
    subtitle = "Property concentration across areas",
    x = "Postcode",
    y = "Count of Properties",
    fill = "Postcode"
  ) +
  theme_minimal()
```

---

## Part 4: Price vs living space by postcode and floor

```{r calc-combos, include=FALSE}
combo_stats <- rental_clean %>%
  filter(!is.na(floor_num), !is.na(postcode_cat)) %>%
  group_by(postcode_cat, floor_num) %>%
  summarise(mean_p = mean(price_chf, na.rm = TRUE), n_count = n(), .groups = "drop") %>%
  filter(n_count >= 1) %>%
  arrange(mean_p)

min_c <- head(combo_stats, 1)
max_c <- tail(combo_stats, 1)
```

```{r plot-complex-scatter}
rental_clean %>%
  filter(!is.na(floor_num)) %>%
  ggplot(aes(x = living_space_m2, y = price_chf, color = postcode_cat)) +
  geom_point() +
  facet_wrap(~ floor_num) +
  labs(
    title = "Price vs Living Space by Floor",
    subtitle = "Faceted by floor level",
    x = "Living Space (m2)",
    y = "Price (CHF)",
    color = "Postcode"
  ) +
  theme_minimal()
```

The lowest average cost combination is postcode `r min_c$postcode_cat` on floor `r min_c$floor_num` (mean: CHF `r round(min_c$mean_p, 2)`). The highest average combination is postcode `r max_c$postcode_cat` on floor `r max_c$floor_num` (mean: CHF `r round(max_c$mean_p, 2)`).

---

# Analysis

## Part 5: Comparing "On Request" vs "Available" addresses

### Price comparison boxplot

```{r boxplot-price}
ggplot(rental_clean, aes(x = address_status, y = price_chf, fill = address_status)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "Price Distribution by Address Status",
    x = "Address Status",
    y = "Price (CHF)",
    fill = "Status"
  ) +
  theme_minimal()
```

### Living space comparison boxplot

```{r boxplot-space}
ggplot(rental_clean, aes(x = address_status, y = living_space_m2, fill = address_status)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "Living Space Distribution by Address Status",
    x = "Address Status",
    y = "Living Space (m2)",
    fill = "Status"
  ) +
  theme_minimal()
```

### Floor level comparison boxplot

```{r boxplot-floor}
ggplot(rental_clean, aes(x = address_status, y = floor_num, fill = address_status)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(
    title = "Floor Level Distribution by Address Status",
    x = "Address Status",
    y = "Floor Number",
    fill = "Status"
  ) +
  theme_minimal()
```

---

## Part 6: Comparing price per square meter

```{r summary-table-m2}
rental_valid_m2 <- rental_clean %>%
  filter(!is.na(price_per_m2), price_per_m2 > 0)

table_m2 <- rental_valid_m2 %>%
  group_by(address_status) %>%
  summarise(
    count = n(),
    median_val = median(price_per_m2),
    mean_val = mean(price_per_m2),
    sd_val = sd(price_per_m2),
    min_val = min(price_per_m2),
    max_val = max(price_per_m2)
  )

kable(table_m2, col.names = c("Status", "N", "Median", "Mean", "SD", "Min", "Max"))
```

### Welch two-sample t-test (CHF/m2)

```{r ttest-m2}
ttest_m2 <- t.test(price_per_m2 ~ address_status, data = rental_valid_m2)
tidy(ttest_m2) %>% kable()
```

---

## Part 7: Comparing total price (CHF)

```{r summary-table-price}
rental_valid_price <- rental_clean %>%
  filter(!is.na(price_chf), price_chf > 0)

table_price <- rental_valid_price %>%
  group_by(address_status) %>%
  summarise(
    count = n(),
    median_val = median(price_chf),
    mean_val = mean(price_chf),
    sd_val = sd(price_chf),
    min_val = min(price_chf),
    max_val = max(price_chf)
  )

kable(table_price, col.names = c("Status", "N", "Median", "Mean", "SD", "Min", "Max"))
```

### Welch two-sample t-test (Total CHF)

```{r ttest-price}
ttest_price <- t.test(price_chf ~ address_status, data = rental_valid_price)
tidy(ttest_price) %>% kable()
```

---

# Mapping sample addresses

## Part 8: Geocoding and Leaflet map

```{r map-addresses}
set.seed(123)

sample_avail <- rental_clean %>%
  filter(address_status == "Available", !is.na(address))

# Sample up to 30 depending on total available count
sample_n_val <- min(30, nrow(sample_avail))

sample_30 <- sample_avail %>%
  slice_sample(n = sample_n_val) %>%
  mutate(full_addr = paste(address, "Switzerland", sep = ", "))

# Geocoding addresses using OpenStreetMap
geocoded <- sample_30 %>%
  geocode(address = full_addr, method = "osm", lat = latitude, long = longitude) %>%
  filter(!is.na(latitude), !is.na(longitude))

# Creating interactive map
leaflet(geocoded) %>%
  addTiles() %>%
  addMarkers(
    lng = ~longitude, 
    lat = ~latitude,
    popup = ~paste0(object_type, "<br>Price: CHF ", price_chf)
  )
```