---
title: "Report 1: Survey Data Analysis"
author: "Igor Donchev"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)
```

In this project, I am analyzing a survey dataset from a conference in the Netherlands. I will load the data, clean up the column names, count some basic stuff like universities and jobs, fix a bad category column, and make a few plots using ggplot2.

# Loading the tools I need

First, I need to load two packages: tidyverse and janitor.

```{r load-packages}
library(tidyverse)
library(janitor)
```

# Loading and Exploring the Data

## Importing survey data from local folder

```{r load-data}
# Mistake 1: Using base R read.csv() instead of tidyverse read_csv()
survey_raw <- read.csv("C:/Users/Валентина Терешкова/Downloads/survey.csv")
```

## Making column names easier to type with clean_names()

```{r clean-names}
survey_clean <- survey_raw %>%
  clean_names()
```

## Checking how many people took the survey

```{r count-people}
total_people <- nrow(survey_clean)
```

## Seeing what kind of data is in each column

```{r inspect-data}
glimpse(survey_clean)
```

## Counting people per university

```{r count-universities}
uni_summary <- survey_clean %>%
  count(university, name = "total_participants") %>%
  arrange(desc(total_participants))
```

## Top 3 schools in a small table

```{r top-3-table}
top_3_unis <- uni_summary %>%
  head(3)

knitr::kable(top_3_unis, col.names = c("University Name", "Number of People"))
```

## Which position appears the most?

I will count the position column to find out what job most participants do.

```{r most-common-job}
most_common_job <- survey_clean %>%
  count(position) %>%
  arrange(desc(n)) %>%
  slice(1) %>%
  pull(position)
```

## Visualizing the job distribution with ggplot2

```{r plot-jobs}
# Mistake 2: Fixed column name typo (position), but removed label rotation, causing overlapping axis labels
survey_clean %>%
  filter(!is.na(position)) %>%
  ggplot(aes(x = position, fill = position)) +
  geom_bar() +
  labs(
    title = "Count of Participants by Position",
    subtitle = "Simple breakdown of jobs in the survey",
    x = "Job Position",
    y = "How Many People",
    fill = "Job Title"
  ) +
  theme_minimal()
```

# Fixing the size bins

## Creating size_group based on number_of_employees

```{r create-size-bins}
# Mistake 3: Overlapping condition order in case_when (<= 50 placed before <= 20)
survey_fixed_bins <- survey_clean %>%
  mutate(
    size_group = case_when(
      number_of_employees <= 50 ~ "Medium (21-50)",
      number_of_employees <= 20 ~ "Small (0-20)",
      number_of_employees <= 200 ~ "Large (51-200)",
      number_of_employees > 200 ~ "Extra Large (>200)"
    ),
    size_group = factor(
      size_group,
      levels = c("Small (0-20)", "Medium (21-50)", "Large (51-200)", "Extra Large (>200)")
    )
  )
```

## Which group size appears most often

```{r most-frequent-size}
top_size_group <- survey_fixed_bins %>%
  count(size_group) %>%
  filter(!is.na(size_group)) %>%
  arrange(desc(n)) %>%
  slice(1)
```

## Make a bar graph showing how many research groups fall into each size bin

```{r plot-size-bins}
survey_fixed_bins %>%
  filter(!is.na(size_group)) %>%
  ggplot(aes(x = size_group, fill = size_group)) +
  geom_bar() +
  labs(
    title = "Research Groups by Size",
    subtitle = "Group size after fixing the category boundaries",
    x = "Group Size Category",
    y = "Number of Research Groups",
    fill = "Category"
  ) +
  theme_minimal()
```

# Calculating percentages for PhDs, Postdocs, and Professors

## Calculating percentages

```{r calculate-percentages}
survey_percents <- survey_fixed_bins %>%
  mutate(person_id = row_number()) %>%
  filter(number_of_employees > 0) %>%
  mutate(
    pct_phd = number_of_ph_d_students / number_of_employees,
    pct_postdoc = number_of_postdocs / number_of_employees,
    pct_prof = number_of_professors / number_of_employees
  ) %>%
  select(person_id, pct_phd, pct_postdoc, pct_prof)
```

## pivot_longer to plot everything together

```{r pivot-longer-roles}
survey_long <- survey_percents %>%
  pivot_longer(
    cols = c(pct_phd, pct_postdoc, pct_prof),
    names_to = "role",
    values_to = "proportion"
  ) %>%
  mutate(
    role = case_when(
      role == "pct_phd" ~ "PhD Students",
      role == "pct_postdoc" ~ "Postdocs",
      role == "pct_prof" ~ "Professors"
    )
  )
```

## Overall average share of PhDs

```{r average-phd-share}
avg_phd <- survey_long %>%
  filter(role == "PhD Students") %>%
  summarise(mean_val = mean(proportion, na.rm = TRUE)) %>%
  pull(mean_val)

avg_phd_percent <- round(avg_phd * 100, 1)
```

## Create a horizontal stacked bar chart for each research group

```{r plot-group-composition}
ggplot(survey_long, aes(x = factor(person_id), y = proportion, fill = role)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Group Composition Breakdown",
    subtitle = "Percentage of PhDs, Postdocs, and Professors per group",
    x = "Group ID",
    y = "Percentage",
    fill = "Role"
  ) +
  theme_minimal()
```