---
title: "Project 1"
author: "Igor Donchev"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)
```

In this file I am trying to grab data from the rental site and clean it up a bit so we can use it later.

# Getting the libraries

I just loaded the packages here because we need them to get the site data and fix the columns.

```{r load-packages}
library(tidyverse)
library(rvest)
library(janitor)
```

# Scraping the site and fixing things

## Part 1a: Getting the stuff from the web page

So here I just put the website link and use rvest to get all the listing boxes from the page and stick them into a table.

```{r scrape-rental-site}
url <- "https://epfl-exts.github.io/rental-scrape/"

page <- read_html(url)
cards <- html_nodes(page, ".col-sm")

rental_raw <- tibble(
  object_type     = html_text(html_node(cards, ".object-type"), trim = TRUE),
  rooms           = html_text(html_node(cards, ".rooms"), trim = TRUE),
  price           = html_text(html_node(cards, ".price"), trim = TRUE),
  address         = html_text(html_node(cards, ".address"), trim = TRUE),
  living_space    = html_text(html_node(cards, ".living-space"), trim = TRUE),
  floor           = html_text(html_node(cards, ".floor"), trim = TRUE),
  usable_surface  = html_text(html_node(cards, ".usable-surface"), trim = TRUE),
  availability    = html_text(html_node(cards, ".availability"), trim = TRUE),
  card_button     = html_text(html_node(cards, ".card-button"), trim = TRUE)
)
```

Let's see how big this table is.

```{r check-dim}
dim(rental_raw)
```

---

## Part 1b: Fixing the numbers and names

The numbers are all mixed up with text right now like "CHF" or "m2", so I'm using `parse_number()` to turn them into real numbers. Also fixing the column names and checking if the address says on request.

```{r clean-rental-data}
rental_clean <- clean_names(rental_raw)

# Pulling out numbers
rental_clean$price_chf <- parse_number(rental_clean$price)
rental_clean$living_space_m2 <- parse_number(rental_clean$living_space)
rental_clean$floor_num <- parse_number(rental_clean$floor)
rental_clean$rooms_num <- parse_number(rental_clean$rooms)
rental_clean$postcode_cat <- as.character(parse_number(rental_clean$address))

# Check if address is hidden
address_lower <- str_to_lower(rental_clean$address)
rental_clean$address_status <- ifelse(
  str_detect(address_lower, "on request") | str_detect(address_lower, "sur demande"),
  "On Request",
  "Available"
)

# Getting price per square meter
rental_clean$price_per_m2 <- rental_clean$price_chf / rental_clean$living_space_m2
```

Let's look at what the dataset looks like now.

```{r inspect-clean}
glimpse(rental_clean)
```

And quick summary of prices and space just to see what we have.

```{r check-ranges}
summary(rental_clean$price_chf)
summary(rental_clean$living_space_m2)
```
## Part 1b: Exploring the data and fixing types

Now I'm checking what data types R gave us for each column and making sure everything is converted properly. I am creating new numeric columns for calculations so we don't lose any of the original scraped information.

```{r check-types}
# First let's see what types we have right now
glimpse(rental_raw)
```

The columns all came in as character strings because they have text like "CHF" or "m2" mixed in. I will extract the numbers out into new numeric columns using `parse_number()` while keeping the original columns intact.

```{r clean-data-types}
rental_clean <- rental_raw %>%
  clean_names() %>%
  mutate(
    # Extracting numeric values without losing raw text
    price_chf       = parse_number(price),
    living_space_m2 = parse_number(living_space),
    floor_num       = parse_number(floor),
    rooms_num       = parse_number(rooms),
    
    # Cleaning up postcode as a clean category code
    postcode_cat    = as.character(parse_number(address)),
    
    # Checking address status
    address_status  = if_else(
      str_detect(str_to_lower(address), "on request|sur demande"),
      "On Request",
      "Available"
    ),
    
    # Calculating price per m2
    price_per_m2    = price_chf / living_space_m2
  )
```

### Adjustments made:
- Kept all 9 original raw character columns intact so no info is lost.
- Converted `price` to numeric `price_chf`.
- Converted `living_space` to numeric `living_space_m2`.
- Converted `floor` to numeric `floor_num`.
- Converted `rooms` to numeric `rooms_num`.
- Created `postcode_cat` by extracting the zip codes from the `address` text.
- Added `address_status` to tag hidden addresses ("On Request" vs "Available").
- Calculated `price_per_m2` for cost analysis.

---

### Checking unique values for character columns

Let's look at all distinct entries in the original text columns to spot any strange values or missing patterns.

```{r check-unique-chars}
# Checking unique categories in object_type and availability
unique(rental_clean$object_type)
unique(rental_clean$availability)
unique(rental_clean$address_status)
```

---

### Checking ranges for numerical variables


```{r check-numeric-ranges}
rental_clean %>%
  select(price_chf, living_space_m2, floor_num, rooms_num, price_per_m2) %>%
  summary()
```

# Visualizing the data

## Part 2: Plotting price vs. living space

I am creating a scatterplot here using `ggplot2` to see if bigger apartments cost more money.

```{r plot-price-space}
library(ggplot2)

ggplot(rental_clean, aes(x = living_space_m2, y = price_chf)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Price vs. Living Space",
    x = "Living Space (m2)",
    y = "Price (CHF)"
  ) +
  theme_minimal()
```

Looking at the plot, there is a upward trend—as living space increases, the price generally goes up as well. 

 The red trend line confirms that bigger places generally cost more money on average.
 
 ## Part 3: Counting properties by postcode

Here I am creating a bar plot using `ggplot2` to count how many rental listings are located in each postcode area. This will help us see if the agency is more active in certain neighborhoods than others.

```{r plot-postcode-bars}
library(ggplot2)

ggplot(rental_clean, aes(x = postcode_cat, fill = postcode_cat)) +
  geom_bar() +
  labs(
    title = "Number of Properties by Postcode",
    x = "Postcode",
    y = "Number of Properties",
    fill = "Postcode"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### My comments

The agency is definitely more active in specific areas
1207 and 1204. On the other hand, outer areas like 1233, 1234, and have very few listings on the site.
The agency concentrates most of its business in key central postcodes rather than spreading properties evenly across all regions.