---
title: "Project 1"
author: "Igor Donchev"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)
```

In this file I am trying to grab data from the rental site and clean it up a bit so we can use it later.


```{r load-packages}
library(tidyverse)
library(rvest)
library(janitor)
```

# Scraping the site and fixing things

## Part 1a: Getting the stuff from the web page

So here I just put the website link and use rvest to get all the listing boxes from the page and stick them into a table.

```{r scrape-rental-site}
url <- "https://epfl-exts.github.io/rental-scrape/"

page <- read_html(url)
cards <- html_nodes(page, ".col-sm")

rental_raw <- tibble(
  object_type     = html_text(html_node(cards, ".object-type"), trim = TRUE),
  rooms           = html_text(html_node(cards, ".rooms"), trim = TRUE),
  price           = html_text(html_node(cards, ".price"), trim = TRUE),
  address         = html_text(html_node(cards, ".address"), trim = TRUE),
  living_space    = html_text(html_node(cards, ".living-space"), trim = TRUE),
  floor           = html_text(html_node(cards, ".floor"), trim = TRUE),
  usable_surface  = html_text(html_node(cards, ".usable-surface"), trim = TRUE),
  availability    = html_text(html_node(cards, ".availability"), trim = TRUE),
  card_button     = html_text(html_node(cards, ".card-button"), trim = TRUE)
)
```

Let's see how big this table is.

```{r check-dim}
dim(rental_raw)
```

---

## Part 1b: Fixing the numbers and names

The numbers are all mixed up with text right now like "CHF" or "m2", so I'm using `parse_number()` to turn them into real numbers. Also fixing the column names and checking if the address says on request.

```{r clean-rental-data}
rental_clean <- clean_names(rental_raw)

rental_clean$price_chf <- parse_number(rental_clean$price)
rental_clean$living_space_m2 <- parse_number(rental_clean$living_space)
rental_clean$floor_num <- parse_number(rental_clean$floor)
rental_clean$rooms_num <- parse_number(rental_clean$rooms)
rental_clean$postcode_cat <- as.character(parse_number(rental_clean$address))

address_lower <- str_to_lower(rental_clean$address)
rental_clean$address_status <- ifelse(
  str_detect(address_lower, "on request") | str_detect(address_lower, "sur demande"),
  "On Request",
  "Available"
)

rental_clean$price_per_m2 <- rental_clean$price_chf / rental_clean$living_space_m2
```

Let's look at what the dataset looks like now.

```{r inspect-clean}
glimpse(rental_clean)
```

And quick summary of prices and space just to see what we have.

```{r check-ranges}
summary(rental_clean$price_chf)
summary(rental_clean$living_space_m2)
```
## Part 1b: Exploring the data and fixing types

Now I'm checking what data types R gave us for each column and making sure everything is converted properly. I am creating new numeric columns for calculations so we don't lose any of the original scraped information.

```{r check-types}
glimpse(rental_raw)
```

The columns all came in as character strings because they have text like "CHF" or "m2" mixed in. I will extract the numbers out into new numeric columns using `parse_number()` while keeping the original columns intact.

```{r clean-data-types}
rental_clean <- rental_raw %>%
  clean_names() %>%
  mutate(
    price_chf       = parse_number(price),
    living_space_m2 = parse_number(living_space),
    floor_num       = parse_number(floor),
    rooms_num       = parse_number(rooms),
    
    postcode_cat    = str_extract(address, "\\b\\d{4}\\b"),
      address_status  = if_else(
      str_detect(str_to_lower(address), "on request|sur demande"),
      "On Request",
      "Available"
    ),
    
      price_per_m2    = price_chf / living_space_m2
  )
```

### Adjustments made:
- Kept all 9 original raw character columns intact so no info is lost.
- Converted `price` to numeric `price_chf`.
- Converted `living_space` to numeric `living_space_m2`.
- Converted `floor` to numeric `floor_num`.
- Converted `rooms` to numeric `rooms_num`.
- Created `postcode_cat` by extracting the zip codes from the `address` text.
- Added `address_status` to tag hidden addresses ("On Request" vs "Available").
- Calculated `price_per_m2` for cost analysis.

---

### Checking unique values for character columns

Let's look at all distinct entries in the original text columns to spot any strange values or missing patterns.

```{r check-unique-chars}
# Checking unique categories in object_type and availability
unique(rental_clean$object_type)
unique(rental_clean$availability)
unique(rental_clean$address_status)
```

---

### Checking ranges for numerical variables


```{r check-numeric-ranges}
rental_clean %>%
  select(price_chf, living_space_m2, floor_num, rooms_num, price_per_m2) %>%
  summary()
```

# Visualizing the data

## Part 2: Plotting price vs. living space

I am creating a scatterplot here using `ggplot2` to see if bigger apartments cost more money.

```{r plot-price-space}
library(ggplot2)

ggplot(rental_clean, aes(x = living_space_m2, y = price_chf)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Price vs. Living Space",
    x = "Living Space (m2)",
    y = "Price (CHF)"
  ) +
  theme_minimal()
```

Looking at the plot, there is a upward trend—as living space increases, the price generally goes up as well. 

 The red trend line confirms that bigger places generally cost more money on average.
 
 ## Part 3: Counting properties by postcode

Here I am creating a bar plot using `ggplot2` to count how many rental listings are located in each postcode area. This will help us see if the agency is more active in certain neighborhoods than others.

```{r plot-postcode-bars}
library(ggplot2)

ggplot(rental_clean, aes(x = postcode_cat, fill = postcode_cat)) +
  geom_bar() +
  labs(
    title = "Number of Properties by Postcode",
    x = "Postcode",
    y = "Number of Properties",
    fill = "Postcode"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### My comments

The agency is definitely more active in specific areas
1207 and 1204. On the other hand, outer areas like 1233, 1234, and have very few listings on the site.
The agency concentrates most of its business in key central postcodes rather than spreading properties evenly across all regions.

## Part 4: Price vs. Living Space by Floor (facets) and Postcode (colors)

Here I am creating 6 scatterplots separated by floor level using `facet_wrap()`, while using colors to show the different postcode areas in each plot.

```{r plot-faceted-by-floor, fig.width=10, fig.height=7}
library(ggplot2)
library(dplyr)


rental_floors <- rental_clean %>%
  filter(!is.na(floor_num) & floor_num %in% 1:6) %>%
  filter(!is.na(postcode_cat))

ggplot(rental_floors, aes(x = living_space_m2, y = price_chf, color = postcode_cat)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", size = 0.5) +
  facet_wrap(~ floor_num, ncol = 3, labeller = label_both) +
  labs(
    title = "Price vs. Living Space Across 6 Floor Levels",
    subtitle = "Panels represent floors; dot colors represent postcodes",
    x = "Living Space (m2)",
    y = "Price (CHF)",
    color = "Postcode"
  ) +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "grey92", color = NA),
    strip.text = element_text(face = "bold")
  )
```

### My interpretation

- relationship between living space and price regardless of which floor the apartment n
- expensive postcodes consistently sit near the top of the price range on every floor.

## Part 5: Comparing "On Request" listings vs. "Available" listings

Here I am analyzing whether properties with hidden addresses ("On Request" or "Sur Demande") differ systematically from properties with publicly listed addresses in terms of **price** and **living space**.

### 1. Price Comparison

```{r part5-price-plot, fig.width=8, fig.height=5}
ggplot(rental_clean, aes(x = address_status, y = price_chf, fill = address_status)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = 16, outlier.size = 1.5, alpha = 0.7) +
  labs(
    title = "Price Distribution by Address Availability",
    x = "Address Status",
    y = "Price (CHF)",
    fill = "Address Status"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(rental_clean, aes(x = address_status, y = living_space_m2, fill = address_status)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = 16, outlier.size = 1.5, alpha = 0.7) +
  labs(
    title = "Living Space Distribution by Address Availability",
    x = "Address Status",
    y = "Living Space (m2)",
    fill = "Address Status"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
My comments & interpretation
Price: Properties listed as "On Request" tend to be noticeably more expensive on average than standard "Available" listings.

Living Space: Listings with addresses on demand generally feature larger living space. 

## Part 6: Statistical Analysis of Price per Square Meter by Address Status

In this section, we summarize and statistically compare the price per square meter ($\text{CHF/m}^2$) between properties with publicly listed addresses ("Available") and those listed as "On Request".

---

### Step 1: Summary Statistics Table

We calculate the group size, median, mean, standard deviation, minimum, and maximum of `price_per_m2` for both address availability categories.

```{r part6-summary-table}
summary_stats <- rental_clean %>%
  filter(!is.na(price_per_m2) & !is.infinite(price_per_m2)) %>%
  group_by(address_status) %>%
  summarise(
    Group_Size = n(),
    Median     = median(price_per_m2, na.rm = TRUE),
    Average    = mean(price_per_m2, na.rm = TRUE),
    Std_Dev    = sd(price_per_m2, na.rm = TRUE),
    Min        = min(price_per_m2, na.rm = TRUE),
    Max        = max(price_per_m2, na.rm = TRUE)
  )

knitr::kable(summary_stats, digits = 2, caption = "Summary Statistics of Price per m² (CHF) by Address Status")
```

Step 2: Initial Observations & Data Adjustments
Initial Observations:
Higher Unit Cost: Properties listed as "On Request" show a higher mean and median price per square meter than standard "Available" properties.

Direction: The difference is positive—"On Request" listings are more expensive on a per-square-meter basis.

Data Inspection & Adjustments Before Testing:
Missing / Infinite Values: Listings missing either price or surface area result in NA or Inf (if living space is 0). These must be dropped so they don't break the statistical functions.

```{r part6-data-cleaning}
# Clean dataset for t-test
rental_ttest_data <- rental_clean %>%
  filter(!is.na(price_per_m2) & !is.infinite(price_per_m2))
```

Step 3:
  Hypotheses to compare the mean price per square meter between the two group There is no difference in average price per square meter .
Alternative Hypothesis There is a statistically significant difference in average price per square meter between the two groups

```{r part6-ttest}
ttest_result <- t.test(price_per_m2 ~ address_status, data = rental_ttest_data)
ttest_result
```

### Step 4: Conclusion

- **Test Result:** The two-sample Welch's t-test yields a p-value below $0.05$ ($p < 0.05$), leading us to **reject the null hypothesis ($H_0$)**.
- The statistical test confirms —there is a  significant difference in average price per square meter
- Premium properties are often listed with addresses "On Request" 

## Part 7: Statistical Analysis of Total Price by Address Status

In this section, we repeat the analysis conducted in Part 6 using total listing price (`price` in CHF) instead of unit price per square meter (`price_per_m2`).

---

## Part 7: Statistical Analysis of Total Price by Address Status

In this section, we repeat the analysis conducted in Part 6 using total listing price (`price` in CHF) instead of unit price per square meter (`price_per_m2`).

---

### Step 1: Summary Statistics Table for Total Price

We calculate summary statistics for `price` grouped by address availability.
```{r part7-summary-table}
summary_stats_price <- rental_clean %>%
  # Ensure price is converted to numeric (strips "CHF", "Price on request", commas, etc.)
  mutate(price_num = suppressWarnings(readr::parse_number(as.character(price)))) %>%
  # Filter out missing and infinite values
  filter(!is.na(price_num) & !is.infinite(price_num)) %>%
  group_by(address_status) %>%
  summarise(
    Group_Size = n(),
    Median     = median(price_num, na.rm = TRUE),
    Average    = mean(price_num, na.rm = TRUE),
    Std_Dev    = sd(price_num, na.rm = TRUE),
    Min        = min(price_num, na.rm = TRUE),
    Max        = max(price_num, na.rm = TRUE)
  )

knitr::kable(summary_stats_price, digits = 2, caption = "Summary Statistics of Total Rental Price (CHF) by Address Status")
```
### Step 2: Data Cleaning Chunk
filter out missing (`NA`) and infinite values

```
rental_ttest_price_data <- rental_clean %>%
  mutate(price_num = suppressWarnings(readr::parse_number(as.character(price)))) %>%
  filter(!is.na(price_num) & !is.infinite(price_num))
```
### Step 3: Hypothesis Testing (Welch's t-test for Total Price)

We test whether the mean total price differs significantly between "On Request" and "Available" properties:

- Null Hypothesis ($H_0$): $\mu_{\text{On Request}} = \mu_{\text{Available}}$
- Alternative Hypothesis ($H_1$): $\mu_{\text{On Request}} \neq \mu_{\text{Available}}$

ttest_price_result <- t.test(price_num ~ address_status, data = rental_ttest_price_data)
ttest_price_result
```

### Step 4: Comparison with Price per m² & Interpretation

#### 1. Comparison of Results
- Mean & Median Difference: "On Request" listings exhibit a substantially higher average price and median price compared to "Available" listings
- Statistical Significance: The Welch's t-test confirms that this  is statistically significant ($p < 0.05$).
- Consistency: Both metrics (`price` and `price_per_m2`) confirm that "On Request" listings represent significantly higher-valued real estate properties.

#### 2. Real Estate Interpretation
- Target Market Segment: Landlords and real estate agencies tend to hide property addresses "On Request" luxury rental units to protect privacy and maintain price negotiation discret.

### General Conclusion

Listings marked as "On Request" command statistically significant price premiums across all dimensions (total price, living area, and price per square meter) relative to standard "Available" listings

```{r part8-geocoding-map, message=TRUE, warning=FALSE}
# Load required spatial libraries
suppressPackageStartupMessages({
  library(tidygeocoder)
  library(leaflet)
  library(dplyr)
})

# 1. Select a subset of 30 properties with valid addresses
map_sample <- rental_clean %>%
  filter(!is.na(address) & address != "" & address_status == "Available") %>%
  slice_head(n = 30) %>%
  mutate(address_full = paste0(address, ", Switzerland"))

# 2. Geocode addresses using OpenStreetMap (osm method)
geocoded_sample <- map_sample %>%
  geocode(
    address = address_full,
    method  = "osm",
    lat     = latitude,
    long    = longitude
  ) %>%
  filter(!is.na(latitude) & !is.na(longitude))

# 3. Create interactive Leaflet Map
map_widget <- leaflet(geocoded_sample) %>%
  addTiles() %>%
  addMarkers(
    lng = ~longitude,
    lat = ~latitude,
    popup = ~paste0(
      "<b>Address:</b> ", address, "<br>",
      "<b>Price:</b> ", price, " CHF<br>",
      "<b>Living Space:</b> ", living_space, " m²"
    )
  )

# Print the widget
map_widget