---
title: "Part 1, 2 & 3 - Survey Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this project I will load the survey database, look at the tables, analyze the answer counts, and plot bar charts to compare student and professor answers.

# Loading Data

First I load the packages.

```{r}
library(DBI)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Now I connect to the database in Downloads.

```{r}
survey_db <- dbConnect(RSQLite::SQLite(), "C:/Users/Валентина Терешкова/Downloads/likert_survey.db")

dbListTables(survey_db)
```

Next I extract the two tables into tibbles and close the connection.

```{r}
professors <- dbReadTable(survey_db, "PROFESSORS") %>% 
  as_tibble()

students <- dbReadTable(survey_db, "STUDENTS") %>% 
  as_tibble()

dbDisconnect(survey_db)
```

# Preview Data

Here is a quick look at the datasets.

```{r}
head(professors)

head(students)
```

# Part 2 - Survey Analysis

## Reshaping Data into Long Format

We use pivot_longer to make the data long so it is easier to count total answers.

```{r}
# Pivot student table
students_long <- students %>% 
  pivot_longer(
    cols = -1, 
    names_to = "question", 
    values_to = "answer"
  ) %>% 
  filter(!is.na(answer))

# Pivot professor table
professors_long <- professors %>% 
  pivot_longer(
    cols = -1, 
    names_to = "question", 
    values_to = "answer"
  ) %>% 
  filter(!is.na(answer))
```

## Total Number of Answers Obtained from Students

```{r}
nrow(students_long)
```

## Total Number of Answers Obtained from Professors

```{r}
nrow(professors_long)
```

## Does Each Question Have the Same Number of Answers?

### Student Answers per Question

```{r}
students_long %>% 
  group_by(question) %>% 
  summarise(n_answers = n())
```

### Professor Answers per Question

```{r}
professors_long %>% 
  group_by(question) %>% 
  summarise(n_answers = n())
```

# Part 3 - Side-by-Side Bar Charts

## Adding Role Labels

First I add a column to each table to know who is a Student and who is a Professor.

```{r}
students_long$role <- "Student"
professors_long$role <- "Professor"
```

## Combining the Datasets

Now I join both tables together into one big table.

```{r}
all_responses <- rbind(students_long, professors_long)
```

## Plotting Bar Charts with Facets

I create bar charts comparing the answers of students and professors, split into small charts for each question using `facet_wrap()`.

```{r}
ggplot(all_responses, aes(x = answer, fill = role)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ question)
```
# Part 3 - Side-by-Side Facet Chart

## Adding Respondent Role

First I add a column so we know who is a student and who is a professor.

```{r}
students_long$role <- "student"
professors_long$role <- "professor"
```

## Combining Tables

Next I combine both tables into one table using `rbind()`.

```{r}
all_responses <- rbind(students_long, professors_long)
```

## Making the Chart

Here is the side-by-side bar chart split by question using `facet_wrap()`.

```{r, fig.width=10, fig.height=8}
ggplot(all_responses, aes(x = answer, fill = role)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ question) +
  labs(
    title = "My side-by-side facet chart",
    subtitle = "Using facet to show each questions",
    x = "type_of_response",
    y = "count_of_response",
    fill = "respondant"
  ) +
  theme(axis.text.x = element_text(angle = 90))
```
# Part 4 - Finding the Most Common Answer (Mode)

In this part, we find the answer that appears most often for each question and for each respondent group (students vs. professors).

```{r}
library(dplyr)
```

## Step 1: Count Answers by Question, Respondent, and Answer Type

First, we group the combined data and count how many times each answer was given.

```{r}
answer_counts <- all_responses %>% 
  group_by(question, role, answer) %>% 
  summarise(answer_count = n()) %>% 
  ungroup()
```

## Step 2: Find the Highest Count for Each Question and Respondent

Now we group by question and role, filter for the maximum count, and rename the columns to match the required output table.

```{r}
most_common_answers <- answer_counts %>% 
  group_by(question, role) %>% 
  filter(answer_count == max(answer_count)) %>% 
  ungroup() %>% 
  rename(
    question_id = question,
    respondent = role,
    answer_type = answer
  )

# Preview the final table
head(most_common_answers)
```
