---
title: "movie_project"
output: html_document
date: "2026-10-07"


## Part 1: API Query Exercises

### 1. Highest-Grossing Dramas from 2010

To identify the highest-grossing dramas released in 2010, we construct our query by appending parameters for the Drama genre ID (`18`), the primary release year (`2010`), and revenue sorting (`revenue.desc`) to our base setup.

```{r}
library(httr)
library(jsonlite)
library(tibble)

api_key <- "f5d15466393ed930c8e0789be72f4cfd"
base_url <- "https://api.themoviedb.org"
target_url <- modify_url(base_url, path = "/3/discover/movie")

query_params_dramas <- list(
  api_key = api_key,
  with_genres = "18",
  primary_release_year = "2010",
  sort_by = "revenue.desc"
)

response_dramas <- GET(target_url, query = query_params_dramas)
data_dramas <- fromJSON(content(response_dramas, "text", encoding = "UTF-8"), flatten = TRUE)

df_dramas <- as_tibble(data_dramas$results)
print(df_dramas)

query_params_coactors <- list(
  api_key = api_key,
  with_cast = "23659,3896"
)

response_coactors <- GET(target_url, query = query_params_coactors)
data_coactors <- fromJSON(content(response_coactors, "text", encoding = "UTF-8"), flatten = TRUE)

df_coactors <- as_tibble(data_coactors$results)
print(df_coactors)
```

The API output successfully retrieved 3 shared films where both Will Ferrell and Liam Neeson are credited:

The Lego Movie (2014)

Daddy's Home 2 (2017)

Anchorman 2: The Legend Continues (2013)



```{r}
query_params_cert <- list(
  api_key = api_key,
  with_cast = "500",
  certification_country = "US",
  certification.lte = "G"
)

response_cert <- GET(target_url, query = query_params_cert)
data_cert <- fromJSON(content(response_cert, "text", encoding = "UTF-8"), flatten = TRUE)

df_cert <- as_tibble(data_cert$results)
print(df_cert)
```

This confirms that Tom Cruise does not have a traditional G-rated family filmography.




##  Finding Pixar's Company ID

To find Pixar's unique identification number, we query the TMDb company search endpoint and extract the ID programmatically from the results data frame.

```{r}
# Query TMDb search/company endpoint for Pixar
search_url <- "https://api.themoviedb.org/3/search/company"
query_params_company <- list(
  api_key = api_key,
  query = "Pixar"
)

response_company <- GET(search_url, query = query_params_company)
data_company <- fromJSON(content(response_company, "text", encoding = "UTF-8"), flatten = TRUE)

# Extract the company ID for Pixar
pixar_id <- data_company$results$id[data_company$results$name == "Pixar"]
print(pixar_id)
```