---
title: "movie_project"
output: html_document
date: "2026-10-07"
---

## Part 1: API Query Exercises

### 1. Highest-Grossing Dramas from 2010

To identify the highest-grossing dramas released in 2010, we construct our query by appending parameters for the Drama genre ID (`18`), the primary release year (`2010`), and revenue sorting (`revenue.desc`).

```{r}
library(httr)
library(jsonlite)
library(tibble)

api_key <- "f5d15466393ed930c8e0789be72f4cfd"
discover_url <- "https://api.themoviedb.org/3/discover/movie"

query_params_dramas <- list(
  api_key = api_key,
  with_genres = "18",
  primary_release_year = "2010",
  sort_by = "revenue.desc"
)

response_dramas <- GET(discover_url, query = query_params_dramas)
data_dramas <- fromJSON(content(response_dramas, "text", encoding = "UTF-8"), flatten = TRUE)

df_dramas <- as_tibble(data_dramas$results)
print(df_dramas)
```

### 2. Tom Cruise Filmography and Certification

Checking Tom Cruise's filmography (ID `500`) against US certification limits:

```{r}
query_params_cert <- list(
  api_key = api_key,
  with_cast = "500",
  certification_country = "US",
  certification.lte = "G"
)

response_cert <- GET(discover_url, query = query_params_cert)
data_cert <- fromJSON(content(response_cert, "text", encoding = "UTF-8"), flatten = TRUE)

df_cert <- as_tibble(data_cert$results)
print(df_cert)
```

**Conclusion:** Running the TMDb query with the US certification filter (`certification.lte = "G"`) surfaces mostly documentation or metadata edge cases rather than traditional G-rated family films.

---

## Finding Pixar’s Company ID

To find Pixar's unique identification number, we query the TMDb company search endpoint and extract the ID programmatically from the results data frame.

```{r}
search_url <- "https://api.themoviedb.org/3/search/company"
res_comp <- GET(search_url, query = list(api_key = api_key, query = "Pixar"))
data_comp <- fromJSON(content(res_comp, "text", encoding = "UTF-8"), flatten = TRUE)

pixar_id <- data_comp$results$id[data_comp$results$name == "Pixar Animation Studios"]
pixar_id
```

The extracted studio ID for Pixar is `3`.

---

## Stage 3: Pixar Filmography & Pagination

In this section, we programmatically query all animated feature films produced by Pixar (Company ID `3`), sort them by descending revenue, and iterate through the API's paginated responses using `purrr::map_dfr()` to build a complete master dataset.

```{r}
library(tidyverse)

query_params_pixar <- list(
  api_key = api_key,
  with_companies = "3",
  sort_by = "revenue.desc",
  page = 1
)

response_pixar <- GET(discover_url, query = query_params_pixar)
data_pixar <- fromJSON(content(response_pixar, "text", encoding = "UTF-8"), flatten = TRUE)

df_pixar_movies <- as_tibble(data_pixar$results)
head(df_pixar_movies)
```

### Handling Pagination with `purrr::map()`

To fetch all available pages dynamically up to the API's total page count limit:

```{r}
total_pages <- data_pixar$total_pages

all_pixar_movies <- map_dfr(1:total_pages, function(p) {
  res <- GET(discover_url, query = list(
    api_key = api_key,
    with_companies = "3",
    sort_by = "revenue.desc",
    page = p
  ))
  parsed <- fromJSON(content(res, "text", encoding = "UTF-8"), flatten = TRUE)
  as_tibble(parsed$results)
})

glimpse(all_pixar_movies)
```


## 4: Comparing Disney and Pixar
To check if Disney's 2006 acquisition of Pixar was a smart move, we look at how well films from both studios scored after 2006. Since movies with only a couple of votes can throw off the average, we only look at films with at least 50 votes.

### 1. Finding Walt Disney Pictures' Company ID
The company search endpoint returns a few different matches for "Disney". We specifically look for "Walt Disney Pictures" to get its exact ID number.

```{r}
search_url <- "https://api.themoviedb.org/3/search/company"
res_disney <- GET(search_url, query = list(api_key = api_key, query = "Walt Disney Pictures"))
data_disney <- fromJSON(content(res_disney, "text", encoding = "UTF-8"), flatten = TRUE)

disney_id <- data_disney$results$id[data_disney$results$name == "Walt Disney Pictures"]
disney_id
```

## 2. Getting Post-2006 Films for Both Studios
To get movies from 2006 onwards, we check the TMDb API documentation for date filters. We use primary_release_date.gte = "2006-01-01" in our query parameters.

Here is the code to pull movies for both Pixar (3) and Disney (2):

```{r}
# Helper function to grab all pages for a studio from 2006 onwards
get_studio_movies <- function(company_id) {
  initial_res <- GET(discover_url, query = list(
    api_key = api_key,
    with_companies = company_id,
    primary_release_date.gte = "2006-01-01",
    page = 1
  ))
  initial_data <- fromJSON(content(initial_res, "text", encoding = "UTF-8"), flatten = TRUE)
  total_p <- initial_data$total_pages
  
  map_dfr(1:total_p, function(p) {
    res <- GET(discover_url, query = list(
      api_key = api_key,
      with_companies = company_id,
      primary_release_date.gte = "2006-01-01",
      page = p
    ))
    parsed <- fromJSON(content(res, "text", encoding = "UTF-8"), flatten = TRUE)
    as_tibble(parsed$results)
  }) %>% mutate(studio = ifelse(company_id == 3, "Pixar", "Disney"))
}

# Pull data for both studios
pixar_2006 <- get_studio_movies(3)
disney_2006 <- get_studio_movies(2)

# Combine them and keep only films with at least 50 votes
combined_movies <- bind_rows(pixar_2006, disney_2006) %>% 
  filter(vote_count >= 50)
```

## 3. Comparing Popularity and Statistical Testing
Now we want to check if Pixar films get better ratings on average than Disney films. Since movies with only a few votes can throw off the average, we filter our data to keep only films with a vote_count of at least 50.

### Comparing Vote Averages with a Boxplot
First, let's plot the vote averages for both studios using a boxplot:

```{r}
library(ggplot2)

ggplot(combined_movies, aes(x = studio, y = vote_average, fill = studio)) +
  geom_boxplot() +
  labs(
    title = "Vote Averages: Pixar vs. Disney (2006+)",
    x = "Studio",
    y = "Vote Average"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Initial remarks
  Looking at the boxplot, Pixar films generally have a higher median vote average compared to Disney films, showing more consistent ratings overall.Statistical Hypothesis and T-TestTo see if the difference we see in the boxplot is real or just random chance, we run a t-test.Null Hypothesis ($H_0$): There is no difference in the average vote scores between Pixar and Disney movies released since 2006.Alternative Hypothesis ($H_a$): Pixar movies have a different average vote score compared to Disney movies.
  
```{r}
t_test_result <- t.test(vote_average ~ studio, data = combined_movies)
t_test_result
```

### Conclusion
Do the test results confirm our observations? Yes, the t-test gives a p-value below 0.05, meaning Pixar's higher average score is statistically significant.

Was it a smart decision by Disney to acquire Pixar? Yes, buying Pixar turned out to be a great strategic move. It brought in a studio that consistently produces highly rated films, helping Disney maintain strong creative and commercial success in modern animation.